---
title: "Cognition: <br>Programming Edition!"
subtitle: "LondonR 2025"
author: "Hannah Frick"
format:
  revealjs: 
    slide-number: true
    footer: <https://www.frick.ws>
    theme: [default, style.scss]
    highlight-style: a11y
    width: 1280
    height: 720
knitr:
  opts_chunk: 
    echo: true
    collapse: true
    comment: "#>"
    fig.path: "figures/"
---


## {.title background-image="images/ken-suarez-4IxPVkFGJGI-unsplash.jpg"}

::: footer
Photo by <a href="https://unsplash.com/@kensuarez?utm_content=creditCopyText&utm_medium=referral&utm_source=unsplash">Ken Suarez</a> on <a href="https://unsplash.com/photos/4IxPVkFGJGI?utm_content=creditCopyText&utm_medium=referral&utm_source=unsplash">Unsplash</a>
:::

:::{.notes}
- a lot of the tidyverse principles are based on ideas from cognitive psychology
- aim is to give you a very basic idea of how our brain works
- to give you a framework through which to assess programming best practices
:::


```{r prep}
#| include: false

library(tidymodels)
```

# Reading and understanding code {.theme-divider-1}

# Research indicates that almost 60% of programmers’ time is spent understanding rather than writing code.

Felienne Hermans - The Programmer's Brain

::: {.notes}
This talk draws heavily from this book. It's great, you might want to read it!
:::

---

## Let's read some code

```{r}
my_fun <- function(a,  
                   b,  
                   c,  
                   d = 2,  
                   e = 3,  
                   g = 22,  
                   j = 324) {  
  i <- a + c  
  w <- (d + e) / ((a + c) + (d + e))  
  four <- e - a  
  f <- four - a  
  ff <- i + w  
  list(f, ff, four)  
}
```

:::{.notes}
- just pseudo code but still:
- so many arguments!
- no speaking names!
- we just make some (very basic) calculations
- why is this thing named four? 
- we return 3 things but can you recall what they are?
- we didn't even use all the arguments!
:::


# Our short-term memory (STM) only holds two to six items.

## This is not even R code!

```{C}
#| code-line-numbers: "|3-7|9-16|10,15|19-26|20,24|"

SEXP Cdqrls(SEXP x, SEXP y, SEXP tol, SEXP chk)
{
    SEXP ans;
    SEXP qr, coefficients, residuals, effects, pivot, qraux;
    int n, ny = 0, p, rank, nprotect = 4, pivoted = 0;
    double rtol = asReal(tol), *work;
    Rboolean check = asLogical(chk);

    ans = getAttrib(x, R_DimSymbol);
    if(check && length(ans) != 2) error(_("'x' is not a matrix"));
    int *dims = INTEGER(ans);
    n = dims[0]; p = dims[1];
    if(n) ny = (int)(XLENGTH(y)/n); /* y :  n x ny, or an n - vector */
    if(check && n * ny != XLENGTH(y))
	error(_("dimensions of 'x' (%d,%d) and 'y' (%d) do not match"),
	      n,p, XLENGTH(y));

    /* These lose attributes, so do after we have extracted dims */
    if (TYPEOF(x) != REALSXP) {
	PROTECT(x = coerceVector(x, REALSXP));
	nprotect++;
    }
    if (TYPEOF(y) != REALSXP) {
	PROTECT(y = coerceVector(y, REALSXP));
	nprotect++;
    }
/* < more code > */
}
```


## This is not even R code!

```{C}
#| code-line-numbers: "3-7"

SEXP Cdqrls(SEXP x, SEXP y, SEXP tol, SEXP chk)
{
    /* define some variables */

    /* check inputs x and y */

    /* protect x and y */

/* < more code > */
}
```

# Our long-term memory (LTM) helps us aggregate items in our STM into chunks.

:::{.notes}
our R knowledge allowed us to make some sense of C code
:::


## What does this do? And how?

```{r}
#| code-line-numbers: "|1,3,4,13"
drop_strata <- function(expr, in_plus = TRUE) {
  if (rlang::is_call(expr, "+", n = 2) && in_plus) {
    lhs <- drop_strata(expr[[2]], in_plus = in_plus)
    rhs <- drop_strata(expr[[3]], in_plus = in_plus)
    if (rlang::is_call(lhs, "strata")) {
      rhs
    } else if (rlang::is_call(rhs, "strata")) {
      lhs
    } else {
      rlang::call2("+", lhs, rhs)
    }
  } else if (rlang::is_call(expr)) {
    expr[-1] <- purrr::map(as.list(expr[-1]), drop_strata, in_plus = FALSE)
    expr
  } else {
    expr
  }
}
```

## Tests show (important) use cases

```{r}
#| eval: false
#| code-line-numbers: "|1,13|3,4,8,9,15,16"

test_that("`drop_strata()` removes strata term in a series of `+` calls", {
  expect_equal(
    drop_strata(rlang::expr(a + strata(x))),
    rlang::expr(a)
  )
  
  expect_equal(
    drop_strata(rlang::expr(a + strata(x) + b)),
    rlang::expr(a + b)
  )
})

test_that("`drop_strata()` does not remove strata in other cases", {
  expect_equal(
    drop_strata(rlang::expr(a * (b + strata(x)))),
    rlang::expr(a * (b + strata(x)))
  )
})
```

## Abstract syntax tree

```{r}
library(lobstr)

ast(strata(x))

ast(a + b)
```

## Abstract syntax tree

```{r}
library(lobstr)

ast(a + strata(x))

ast(a + strata(x) + b)
```

:::{.notes}
- we go up and down that tree structure
- recursive: which level are we at? -> keep track of that
- keep track of what expr is
:::



# Our working memory is our STM applied to a problem.
# Our working memory only holds two to six items.


## Challenges 

::: columns
::: {.column width="50%"}
- Lack of information
- Lack of knowledge
- Lack of processing power
:::
::: {.column width="50%"}
:::
:::

## Challenges, for reasons

::: columns
::: {.column width="50%"}
- Lack of information
- Lack of knowledge
- Lack of processing power
:::
::: {.column width="50%"}
- Limited capacity of STM
- Activation of LTM
- Limited capacity of working memory
:::
:::


## Help your brain out

. . .

- Look for beacons: names, comments, paragraphs
- Summarize code into chunks via comments or refactoring

. . .

- Learn more: programming concepts, domain knowledge

. . .

- Offload information



# Writing Code {.theme-divider-1}

<!-- the empty notes block is so that the text is in the same place as on the next slide -->
:::{.notes}
:::

# Writing is for re-reading

:::{.notes}
- optimize for the real bottleneck: us humans
:::

# Well-scoped code supports chunking.

# Names {.theme-divider-2}

# Good names help activate knowledge from your LTM.

## {.center}

::: r-fit-text
```
predict(survival_model, 
        type = "survival",  
        time = 2)
```
:::

. . .

`time`: the time points at which the survival probability is estimated


# Bad names can hinder you by activating the wrong knowledge.

## {.center}

::: r-fit-text
```
predict(survival_model, 
        type = "survival",  
        eval_time = 2)
```
:::

# Make (re)thinking names a separate step

to avoid overloading your working memory.

# Bad names are linguistic anti-patterns, code smells are structural anti-patterns.

# Design patterns {.theme-divider-2}

# Design patterns are reusable solutions to common problems.

:::{.notes}
- Learn once, apply many times
- Learn once, recognize many times
:::

--- 

## Many arguments

```{r}
my_fun <- function(x, 
                   y,
                   opt1 = 1,
                   opt2 = 2, 
                   opt3 = 3, 
                   opt4 = 4){
  ...  
}
```

:::{.notes}
- tidymodels: `tune_grid()`
- options: `verbose`, `save_pred`, `extract`, parallel processing
:::


## Many arguments, revisited

```{r}
my_fun <- function(x, y, options = my_fun_opts()) {
  ...
}

my_fun_opts <- function(opt1 = 1, opt2 = 2, opt3 = 3, opt4 = 4) {
  list(
    opt1 = opt1,
    opt2 = opt2,
    opt3 = opt3, 
    opt4 = opt4
  ) 
}
```

:::{.notes}
- tidymodels: `control_grid()`, 
- also: `control_bayes()`, `control_parsnip()`
:::


# Design patterns can help lower the cognitive load.


<!-- ## [outro] -->

##  {.title background-image="images/ken-suarez-4IxPVkFGJGI-unsplash.jpg"}

::: footer
Photo by <a href="https://unsplash.com/@kensuarez?utm_content=creditCopyText&utm_medium=referral&utm_source=unsplash">Ken Suarez</a> on <a href="https://unsplash.com/photos/4IxPVkFGJGI?utm_content=creditCopyText&utm_medium=referral&utm_source=unsplash">Unsplash</a>
:::

:::{.notes}
- want to leave you with two things
:::


# Having a mental model of how your brain works helps you to [work _with_ it, not against it]{.yellow}.

:::{.notes}
- knowing how the limits manifest, knowing how to support yourself
- your mind can only hold so much at a time
- support it so you can focus that capacity on what matters to you
:::

# If you want your code to grow in complexity, <br>you need to keep (re-)chunking. 

:::{.notes}
- why refactoring and upkeep matter
:::

# Keynote at LatinR 2023

Full talk at <https://hfrick.github.io/2023-latinr/>
